TMR: Detecting Structure in Graphical Data

### 2021-01-19 interim-report/interim-report.tex
Cut down to 6

Though the definition of the SBM is simple, it allows for very deep and rich analysis of graphical datasets. For certain problems it helps to define the symmetric SBM.

\begin{definition}
	\label{defn:sym-sbm}
	The symmetric SBM is a special case denoted by $\textrm{SSBM}(n, k, q_{in}, q_{out}) \equiv \textrm{SBM}(n, p, W)$ if the community prior $p$ is uniform ($p_i = 1/k$ for $i \in \{1, 2, \dots k\}$) and $W_{ij}$ takes only two values, one on diagonal and another off diagonal such that $W_{ij} = q_{in}$ for $i=j$ and $W_{ij} = q_{out}$ for $i \neq j$.
\end{definition}

@@

Through application of the Central Limit Theorem (CLT) and Slutsky's Theorem it can be shown that as $n \rightarrow \infty$ under the null $H_0$ the term in brackets on the final line is distributed like a standard Gaussian $z_n \sim \Gaussian(0,1)$. This is because under the same assumptions, assuming that $W_\alpha = W_\beta = \mu$, by the CLT $\hat{p} \sim (\mu, \mu(1 - \mu)/ M_\alpha)$, $\hat{q} \sim (\mu, \mu(1 - \mu)/ M_\beta)$ and $\hat{r} \sim (\mu, \mu(1 - \mu)/ (M_\alpha + M_\beta))$. The proof is long-winded and therefore omitted from this short report. Therefore, the whole test-statistic is distributed as $t_n \sim \frac{1}{2} \chi^2_1$ by the definition of the $\chi^2_p$ distribution: $\chi^2_p \sim \sum_{i=1}{p} Z_i^2$ for $Z_i \sim \Gaussian(0,1)$. This leads to theorem \ref{theorem:hyp-test-sbm-chi} and corollary \ref{corollary:z-test}.

@@

\begin{corollary}[z-test]
	We can also use a slightly simpler test-statistic $z_n$ albeit with some loss of generality. If we define $z_n$ to be:
	%
	\begin{equation*}
		z_n \coloneqq \frac{\hat{p} - \hat{q}}{\sqrt{\hat{r}(1- \hat{r})(1/M_\alpha + 1/M_\beta)}}
	\end{equation*}

	With the symbols retaining their previous meanings. Under the null $H_0$, as $n \rightarrow \infty$ then $z_n \sim \Gaussian(0, 1)$. Meaning that, we can construct a similar test to reject $H_0$ at the $100(\zeta)\%$ confidence level if and only if $|z_n| \geq \phi^{-1}(\zeta)$, where $\phi^{-1}$ is the standard inverse Gaussian cdf on magnitude satisfying $Pr(|Y| \leq \phi^{-1}(\zeta)) = \zeta$ given $Y \sim \Gaussian(0, 1)$
	\label{corollary:z-test}
\end{corollary}

@@

Test 2 rejecting the null gives evidence that gender-78 people (community 2) have on average more same-gender friends than 77-ers have ($\hat{p_2} > \hat{p_1}$ in table \ref{tab:egonet-0-props}). The rejection of the null in test 3 gives evidence that gender-78 treat gender-77 people differently to their own gender. Indeed, a gender-78 person is more likely to be friends with a fellow 78-er than with a 77-er. However, we do not reject the null in test 1 so there is insufficient evidence to claim the reverse (that 77-ers tend to stay away from 78-ers). Nevertheless, these are results on a single egonet so can hardly be generalised to society as a whole. However, the exercise has highlighted some interesting points:

@@

This is the problem of recovering the communities from the graph $\Gcal$. We use the notation $\hat{X}$ to denote the produced estimate of the community labels for the $n$ vertices. Obviously, if the SBM is symmetric then we must allow for an arbitrary relabelling of the communities $r: \{1, 2 \dots k\} \mapsto \{1, 2 \dots k\}$ before we compare the agreement of the two vectors $\hat{X}$ and $X$. The agreement $A$ and normalised agreement $\tilde{A}$ between two vectors $x, y$ are computed as below:
%
\begin{align}
	A(x, y) &= \max_r \frac{1}{n} \sum_{i=1}^{n} \one (x_i, r(y_i)) \\
	\tilde{A}(x, y) &= \max_r \frac{1}{k} \sum_{i=1}^{k} \frac{1}{n_i} \sum_{v \in \Omega_i(x)} \one(x_v, r(y_v))
\end{align}

The normalised agreement is important for asymmetric SBMs as it takes the agreement averaged over communities rather than vertices. We define 4 recovery regimes (plus the trivial no recovery), such that the following holds (adapted from Abbe \cite{Abbe}):

\begin{definition}[Recovery regimes]
	for $(X, \Gcal) \sim \textrm{SBM}(m, p, W)$ the following recovery requirements are solved if there exists an algorithm which takes $\Gcal$ and input and outputs a vector of classifications $\hat{X} = \hat{X}(\Gcal)$ such that as $n \rightarrow \infty$:
	\begin{itemize}
		\item \textbf{Exact recovery:} $Pr\{A(X, \hat{X}) = 1\} = 1 - o(1)$ correct classification recovered almost surely
		\item \textbf{Almost exact recovery:} $Pr\{A(X, \hat{X}) = 1 - o(1)\} = 1 - o(1)$ vanishing fraction misclassified almost surely
		\item \textbf{Partial recovery:} $Pr\{\tilde{A}(X, \hat{X}) \geq \alpha\} = 1 - o(1), 1 > \alpha > 1/k$ better than choosing from uniform prior
		\item \textbf{Weak recovery:} $Pr\{\tilde{A}(X, \hat{X}) \geq 1/k + \epsilon \} = 1 - o(1), \epsilon > 0$ marginally better than choosing from uniform prior
		\item \textbf{No recovery:} graph provides no information as to class labels
	\end{itemize}
	Where $o(1)$ denotes a family of functions that tend to 0 as $n \rightarrow \infty$. The regime $\Gcal$ occupies depends on the parameters of the SBM. Each recovery regime is weaker than the one above it so Exact recovery implies weak recovery is possible..
\end{definition}

We choose to focus on weak recovery as it is the least strict of all regimes. An alternative statement of weak recovery that implies the one above is given below \cite{Abbe}:

\begin{definition}[Weak recovery]
	\label{defn:weak-recovery}
	for  $(X, \Gcal) \sim \textrm{SBM}(n, p, W)$ weak recovery is solved if there exists $\epsilon > 0$ and a certain choice of indices $i, j \in \{1, 2 \dots k\}$ and an algorithm that takes as input $\Gcal$ and outputs a partition of the vertex set $\Vcal$ into two distinct sets $S$ and $S^C$ (detected communities 1 and 2 respectively) such that:
	\begin{equation*}
		Pr\left\{ \frac{|\Omega_i \cap S|}{|\Omega_i|} - \frac{|\Omega_j \cap S|}{|\Omega_j|} \geq \epsilon \right\} = 1 - o(1)
	\end{equation*}

@@

%\nocite{*}
\printbibliography

>>>>

Though the definition of the SBM is simple, it allows for very deep and rich analysis of graphical datasets.

@@

Through application of the Central Limit Theorem (CLT) and Slutsky's Theorem it can be shown that as $n \rightarrow \infty$ under the null $H_0$ the term in brackets on the final line is distributed like a standard Gaussian $z_n \sim \Gaussian(0,1)$. This is because under the same assumptions, assuming that $W_\alpha = W_\beta = \mu$, by the CLT $\hat{p} \sim (\mu, \mu(1 - \mu)/ M_\alpha)$, $\hat{q} \sim (\mu, \mu(1 - \mu)/ M_\beta)$ and $\hat{r} \sim (\mu, \mu(1 - \mu)/ (M_\alpha + M_\beta))$. The proof is long-winded and therefore omitted from this short report. The result leads to theorem \ref{theorem:hyp-test-sbm-chi}.

@@

Note that an alternative test is possible by comparing $z_n$ to a Gaussian.

@@

Test 2 rejecting the null gives evidence that gender-78 people (community 2) have on average more same-gender friends than 77-ers have ($\hat{p}_2 > \hat{p}_1$ in table \ref{tab:egonet-0-props}). The rejection of the null in test 3 gives evidence that gender-78 treat gender-77 people differently to their own gender. Indeed, a gender-78 person is more likely to be friends with a fellow 78-er than with a 77-er. However, we do not reject the null in test 1 so there is insufficient evidence to claim the reverse (that 77-ers tend to stay away from 78-ers). Nevertheless, these are results on a single egonet so can hardly be generalised to society as a whole. However, the exercise has highlighted some interesting points:

@@

Community detection is the problem of recovering the communities from the graph $\Gcal$. We use the notation $\hat{X}$ to denote the estimate of the community labels for the $n$ vertices. Obviously, if the SBM is symmetric then we must allow for an arbitrary relabelling of the communities $r: \{1, 2 \dots k\} \mapsto \{1, 2 \dots k\}$ before we compare the agreement of the two vectors $\hat{X}$ and $X$. There are 4 recovery regimes (plus the trivial no recovery), as defined by Abbe \cite{Abbe}. We choose to focus on weak recovery as it is the least strict of all. Therefore, there are very few constraints on the SBM parameters and the theory can be applied to a wide range of graphs. A statement of weak recovery is given in definition \ref{defn:weak-recovery} \cite{Abbe}.

\begin{definition}[Weak recovery]
	\label{defn:weak-recovery}
	weak recovery is solved for a graph  $(X, \Gcal) \sim \textrm{SBM}(n, p, W)$ if there exists $\epsilon > 0$ and a certain choice of indices $i, j \in \{1, 2 \dots k\}$ and an algorithm that takes as input $\Gcal$ and outputs a partition of the vertex set $\Vcal$ into two distinct sets $S$ and $S^C$ (detected communities 1 and 2 respectively) such that:
	\begin{equation*}
		Pr\left\{ \frac{|\Omega_i \cap S|}{|\Omega_i|} - \frac{|\Omega_j \cap S|}{|\Omega_j|} \geq \epsilon \right\} = 1 - o(1)
	\end{equation*}

@@

\printbibliography
